Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-20 _tex/paper2.tex
Built site for gh-pages

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Therefore, it is reasonable to assume that the outcome's ability to
reflect the ``true'' differences between stimuli largely depends on the
validity of the model's assumptions, particularly the assumption of
equal dispersions. For instance, when researchers observe a sample of
outcomes favoring Text B over Text A and correctly assume equal
dispersions between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated in
Figure~\ref{fig-dispersion}, where the model's discriminal difference
distribution aligns with the ``true'' distribution, represented by the
thick continuous line corresponding to \(\sigma_{B}-\sigma_{A}=0\). The
accuracy of these discriminal difference ensures reliable estimates for
the texts' discriminal processes {(citation needed?)}.

@@

However, in practice, researchers approach this process in reverse. They
begin by observing the sample of outcomes favoring Text B over Text A
and then use the BTL model to estimate the discriminal difference and
the discriminal processes of the stimuli. Given that the BTL model
assumes independent discriminal processes across comparisons, if this
assumption holds, then the model estimates a discriminal difference
distribution that accurately reflects the ``true'' discriminal
difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

@@

This section introduces a theoretical model for Comparative Judgment
(CJ) that builds upon Thurstone's theory. The model systematically
accounts for the factors influencing judges during CJ experiments. These
theoretical and practical insights are then translated into a
statistical model.

@@

\paragraph{Figure 3}

@@

\paragraph{Figure 4}

@@

\paragraph{Figure 5}

@@

\paragraph{Figure 6}

@@

\paragraph{Figure 7}

@@

\paragraph{Figure 8}

@@

\paragraph{Figure 9}

@@

\paragraph{Figure 10}

@@

\paragraph{Figure 11}

@@

\paragraph{Figure 12}

@@

Considering the sampling process

\paragraph{Figure 13}

@@

\paragraph{Figure 14}

@@

\paragraph{Figure 15}

>>>>

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the DP for any given stimulus
forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the mDP for Text B is positioned
further along the continuum than that of Text A \((T_{B} > T_{A})\),
suggesting that Text B exhibits higher quality. Additionally, the figure
highlights that Text B has a broader distribution compared to Text A,
which arises from its larger discriminal dispersion
\((\sigma_{B} > \sigma_{A})\).

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Therefore, it is reasonable to assume that the outcome's ability to
reflect the ``true'' differences between stimuli largely depends on the
validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
particularly the assumption of equal dispersions. For instance, when
researchers observe a sample of outcomes favoring Text B over Text A and
correctly assume equal dispersions between the texts, the BTL model
estimates a discriminal difference distribution that accurately
represents the ``true'' discriminal difference of the texts. This
scenario is illustrated in Figure~\ref{fig-dispersion}, where the
model's discriminal difference distribution aligns with the ``true''
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
difference ensures reliable estimates for the texts' discriminal
processes {(citation needed?)}.

@@

Off course, in experimental practice, researchers approach this process
in reverse. They begin by observing the sample of outcomes favoring Text
B over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

@@

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal analysis
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014, Hernan_et_al_2020}
and, in particular, directed acyclic graphs (DAGs)
\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
principles of CJ theory. The model also incorporates several practical
factors that influence judges in CJ experiments, such as the selection
of judges, stimuli, and the number of comparisons. In addition, the
study uses Bayesian statistical methods \citep{McElreath_2020} to
translate these theoretical and practical elements into a statistical
model that facilitates the analysis of pairwise comparison data.

@@

\paragraph{Figure}

\paragraph{3}

@@

\paragraph{4}

@@

\paragraph{5}

@@

\paragraph{6}

@@

\paragraph{Figure}

\paragraph{7}

@@

\paragraph{8}

@@

\paragraph{9}

@@

\paragraph{10}

@@

\paragraph{11}

@@

\paragraph{12}

@@

Considering the sampling and comparison mechanisms

\paragraph{Figure}

\paragraph{13}

@@

\paragraph{14}

@@

\paragraph{15}
